---
title: "IKDC Questionnaire Scoring"
output:   
  github_document:
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document provides an example of the IKDC questionnaire scoring process. The IKDC is a patient-reported measure of knee-specific symptoms, function, and sports activity. Scores range from 0 to 100, with higher scores indicating higher levels of function.

```{r import, message = FALSE}
data <- read.csv("ikdc_example_sml.csv", stringsAsFactors = FALSE)
```

```{r head}
str(data)
```
In this example, the responses are stored as text strings. The text responses need to be converted into item-level scores. Some items have equivalent scoring - we'll create functions to score those items.

```{r functions}
# Recode item response: "unable to do" = 0 / "not difficult at all" = 4 
ik_unable_0 <- function(var) {
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "unable to do"] <- 0
  res[var == "extremely difficult"] <- 1
  res[var == "moderately difficult"] <- 2
  res[var == "minimally difficult"] <- 3
  res[var == "not difficult at all"] <- 4
  return(res)
}

# Recode item response: "zero" = 10 / "ten" = 0
ik_numeric_rev <- function(var) {
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "ten"] <- 0
  res[var == "nine"] <- 1
  res[var == "eight"] <- 2
  res[var == "seven"] <- 3
  res[var == "six"] <- 4
  res[var == "five"] <- 5
  res[var == "four"] <- 6
  res[var == "three"] <- 7
  res[var == "two"] <- 8
  res[var == "one"] <- 9
  res[var == "zero"] <- 10
  return(res)
} 
```

The following scoring functions are only used once (and thus aren't necessary to create -  but it's nice to have all the recoding information in one place).

``` {r functions misc}
ik_crfunc <- function(var) {
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "ten"] <- 10
  res[var == "nine"] <- 9
  res[var == "eight"] <- 8
  res[var == "seven"] <- 7
  res[var == "six"] <- 6
  res[var == "five"] <- 5
  res[var == "four"] <- 4
  res[var == "three"] <- 3
  res[var == "two"] <- 2
  res[var == "one"] <- 1
  res[var == "zero"] <- 0
  return(res)
} 

ik_perf <- function(var){
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "very strenuous activities like jumping or pivoting as in basketball or soccer"] <- 4
  res[var == "strenuous activities like heavy physical work, skiing or tennis"] <- 3
  res[var == "moderate activities like moderate physical work, running or jogging"] <- 2
  res[var == "light activities like walking, housework, or yard work"] <- 1
  res[var == "unable to perform any of the above activities due to knee pain"] <- 0
  return(res)
}

ik_stiff <- function(var){
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "not at all"] <- 4
  res[var == "mildly"] <- 3
  res[var == "moderately"] <- 2
  res[var == "very"] <- 1
  res[var == "extremely"] <- 0
  return(res)
}

ik_swell <- function(var){
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "very strenuous activities like jumping or pivoting as in basketball or soccer"] <- 4
  res[var == "strenuous activities like heavy physical work, skiing or tennis"] <- 3
  res[var == "moderate activities like moderate physical work, running or jogging"] <- 2
  res[var == "light activities like walking, housework, or yard work"] <- 1
  res[var == "unable to perform any of the above activities due to knee swelling"] <- 0
  return(res)
}

ik_lock <- function(var){
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "no"] <- 1
  res[var == "yes"] <- 0
  return(res)
}

ik_give <- function(var){
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "very strenuous activities like jumping or pivoting as in basketball or soccer"] <- 4
  res[var == "strenuous activities like heavy physical work, skiing or tennis"] <- 3
  res[var == "moderate activities like moderate physical work, running or jogging"] <- 2
  res[var == "light activities like walking, housework, or yard work"] <- 1
  res[var == "unable to perform any of the above activities due to giving way of the knee"] <- 0
  return(res)
}

ik_partic <- function(var){
  res <- rep(999, length(var))
  res[is.na(var)] <- NA
  res[var == "very strenuous activities like jumping or pivoting as in basketball or soccer"] <- 4
  res[var == "strenuous activities like heavy physical work, skiing or tennis"] <- 3
  res[var == "moderate activities like moderate physical work, running or jogging"] <- 2
  res[var == "light activities like walking, housework, or yard work"] <- 1
  res[var == "unable to perform any of the above activities due to knee"] <- 0
  return(res)
}
```

We can now recode the text responses into item-level scores.

```{r item scoring}
scores <- within(data, {
  # Variables that can be recoded with the ik_numeric_rev function
  PAIN <- ik_numeric_rev(PAIN)
  SEV <- ik_numeric_rev(SEV)
  # Variables that can be recoded with the ik_unable_0 function
  UPSTRS <- ik_unable_0(UPSTRS)
  DNSTRS<- ik_unable_0(DNSTRS)
  KNEEL<- ik_unable_0(KNEEL)
  SQUAT<- ik_unable_0(SQUAT)
  SIT<- ik_unable_0(SIT)
  RISE<- ik_unable_0(RISE)
  RUN<- ik_unable_0(RUN)
  JUMP<- ik_unable_0(JUMP)
  STOP<- ik_unable_0(STOP)
  # Variables with unique coding schemes
  CRFUNC <- ik_crfunc(CRFUNC)
  PERF<-ik_perf(PERF)
  STIFF<-ik_stiff(STIFF)
  SWELL<-ik_swell(SWELL)
  LOCK<-ik_lock(LOCK)
  GIVE<-ik_give(GIVE)
  PARTIC<-ik_partic(PARTIC)
})

```

It can be useful to store information about a dataset in a list object. This meta-data can then be used programmatically.

`dput(names(data_frame))` is a useful command for getting lists of variable names when you don't want to use column positions (e.g., `data_frame[,1]`). This will print all variable names into the console - including quotation marks and commas. This text can then easily be copied and pasted into an R script.
```{r dput}
dput(names(data))
```

To score the IKDC, we will eventually need to perform different calculations based on the maximum value of each item. The variable names for items with a maximum value of 1 are stored in `var$max_1`, items with a maximum value of 4 are stored in `var$max_4`, and items with a maximum value of 10 are stored in `var$max_10`. `var$varlist` is a vector of all questionnaire item variable names. We can add infromation to this list as needed.

```{r q-info}
var <- list()
var$max_1 <- c("LOCK")
var$max_4 <- c("PERF","STIFF","SWELL","GIVE","PARTIC","UPSTRS",
                 "DNSTRS","KNEEL","SQUAT","SIT","RISE","RUN","JUMP","STOP")
var$max_10 <- c("PAIN","SEV", "CRFUNC")
var$varlist <- c(var$max_1, var$max_4, var$max_10)
```

The IKDC can only be scored if participants repond to 16 items or more. We'll record a count of the number of non-missing responses to be used later. Only columns that appear in the variable list will be included (i.e., the `ID` field is ignored).

``` {r na count}
scores$response_count <- rowSums( !is.na( scores[ colnames(scores) %in% var$varlist ]))
```

The IKDC score is calculated by: [sum of the item level scores] / [sum of the maximum scores possible on the items to which the participants responded] * 100. 

IKDC items have maximum scores of either 1, 4, or 10. To prepare for calculating the denominator value, we'll create a version of each IKDC item that contains either: NA (if the participant did not respond) or the maximum possible value (if the participant did respond). These variables will have the same names as the item-level score variables, but with a "_denom" suffix.

```{r max-vals}
# Add a denominator suffix ("_denom") to all questionnaire item names in the meta-data list object
var$denom_1 <- paste0(var$max_1, "_denom")
var$denom_4 <- paste0(var$max_4, "_denom")
var$denom_10 <- paste0(var$max_10, "_denom")
var$denom <- paste0(var$varlist, "_denom")

# Recode items based on maximum value
scores[var$denom_10] <- 10
scores[var$denom_4] <- 4
scores[var$denom_1] <- 1
scores[var$denom][is.na(scores[var$varlist])] <- NA

# This works because we know the order of variables in var$denom and var$varlist are the same.
# i.e., first LOCK / LOCK_denom, then PERF / PERF_denom, etc.

```

Now we can calculate the IKDC score. The numerator is created by summing the item-level scores. The denominator is created by summing the maximum scores for non-missing items.

```{r scores}
scores$numerator   <- rowSums(scores[,var$varlist], na.rm=TRUE)
scores$denominator <- rowSums(scores[,var$denom], na.rm=TRUE)
scores$ikdc_score  <- ifelse(scores$response_count > 15, (scores$numerator/scores$denominator)*100, NA)
```

The scores can be merged back into the original dataset.

```{r merge}
data <- merge(data, scores[c("ID", "ikdc_score")], by = "ID", all=TRUE)
```

IKDC scores range from 0 to 100. Let's look at a summary:
```{r summary}
library(dplyr)

data %>% summarize(
  min = min(ikdc_score, na.rm=TRUE),
  mean = mean(ikdc_score, na.rm=TRUE),
  max = max(ikdc_score, na.rm=TRUE),
  missing = sum(is.na(ikdc_score))
  )
```

We can also easily see the score's distribution.
```{r histogram}
library(ggplot2)
theme_set(theme_light())

data %>% 
  filter(!is.na(ikdc_score)) %>% 
  ggplot(aes(ikdc_score))+
  geom_histogram(bins=40) +
  expand_limits(x = 0, y = 0) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +
  geom_vline(aes(xintercept=mean(ikdc_score)), 
             color="red", linetype="dashed", size=1)
```